### 2018-07-13 submission/manuscript.Rmd
revamp results section

@@

The median error rate varied by polymerase, with Kappa having the highest error rate of all the polymerases across the number of cycles [Figure 2A & Table S2]. Thus, the majority of the differences observed across the number of cycles was between Kappa and other polymerases [Figure 2A & Table S3]. The total sequences that contained at least one error was also polymerase dependent, with the majority of differences being between Kappa or Accuprime and the other polymerases [Table S2 & S3]. These differences in error rates were not due to polymerase dependent differences in base substitution rate [Figure S1]. Collectively, the results suggest that sequence error is dependent on polymerase, persists across the number of cycles used, and are not due to any bias towards a specific base substitution.

Based on the previous results, we examined whether chimeras also were dependent on polymerase and whether this could affect the total number of OTUs. We observed significant differences in the chimera prevalence based on polymerase across the number of cycles used (P-value `r ifelse(max(filter(kruskal_chimera_prev)[, "bh"]) < 0.05, invisible("<"), invisible("check data"))` 0.05) [Table S2]. Differences in chimera prevalence between Platinum and all other polymerases accounted for the majority of these differences [Table S3]. Accuprime^TM^ had the lowest chimera prevalence of all polymerases regardless of whether chimera removal with VSEARCH was used [Figure 2B & 2C]. The number of cycles used clearly increased chimera prevalence for all polymerases but the rate of increase differed [Figure 2B & 2C]. Additionally, there was a plateau in the total percent of chimeras that were removed that was similar for all polymerases [Figure 2D]. A positive correlation was observed between chimeric sequences and the number of OTUs for all polymerases [Figure 2E]. This positive correlation was strongest for Accuprime^TM^, Platinum, and Phusion [Figure 2E]. This data suggests that chimera prevalence depends on polymerase, is made worse by increasing the number of cycles, and confirms that the number of OTUs is dependent on the prevalence of these chimeric sequences.

@@

Next, we investigated the effect polymerases and number of cycles had on the bacterial community composition in more detail by using the Bray-Curtis index. We observed no difference in the bacterial community composition between any measure tested when using the mock community (P-value `r ifelse(pull(mock_kruskal_test_by_cycle_compare_summary, p.value) < 0.05, invisible("<"), invisible(">"))` 0.05) [Figure 4A]. Finding minimal differences in overall bacterial community composition, we next examined if specific taxa vary depending on polymerase and number of cycles. We found that OTUs that taxonomically classified to *Salmonella*, *Escherichia*, *Enterococcus*, and *Staphylococcus* differed based on polymerase used [Table S6 & S7]. These specific OTUs were all members of the mock community and the differences were most pronounced at 25 and 30 cycles [Table S7]. The majority of the differences between taxa were due to differences between a single polymerase (Kappa) and the other polymerases [Table S7].

When assessing the Random Forest models there was a difference between models built to classify number of cycles versus those built to classify polymerase used [P-value = `r format(filter(dunn_rf_data, str_detect(var1, "mock") & str_detect(var2, "mock")) %>% pull(bh), scientific = T, digits = 3)`]. Although there was a difference, neither model was successful at correctly classifying samples based on 16S rRNA gene sequencing data (cycles model, probability of correct classification = `r round(exp(-median(mock_cycles_rf_summary$logLoss)), digits = 2)` (`r round(exp(-max(mock_cycles_rf_summary$logLoss)), digits = 2)` - `r round(exp(-min(mock_cycles_rf_summary$logLoss)), digits = 2)`)(min - max) versus polymerase model, probability of correct classification = `r round(exp(-median(mock_poly_rf_summary$logLoss)), digits = 2)` (`r round(exp(-max(mock_poly_rf_summary$logLoss)), digits = 2)` - `r round(exp(-min(mock_poly_rf_summary$logLoss)), digits = 2)`)). Our observations in mock communities suggest that although differences are detected based on number of cycles or polymerase used, the overall community-wide differences are subtle.

***Few differences based on number of cycles and polymerased used were identified in fecal samples.***
Similarly, differences in the range of the number of OTUs detected for fecal samples was dependent on the polymerase used (e.g. Accuprime at 35 cycles = `r filter(fecal_numOTU_data, cycles == "35x", taq == "Accuprime") %>% pull(taq_min)` - `r filter(fecal_numOTU_data, cycles == "35x", taq == "Accuprime") %>% pull(taq_max)` versus Phusion at 35 cycles = `r filter(fecal_numOTU_data, cycles == "35x", taq == "Phusion") %>% pull(taq_min)` - `r filter(fecal_numOTU_data, cycles == "35x", taq == "Phusion") %>% pull(taq_max)`) [Figure 3B]. There was also a trend for lower number of cycles (15-20) to result in a reduced range in the number of OTUs versus higher number of cycles (25, 30, and 35) for all polymerases (e.g. Phusion at 15 cycles = `r filter(fecal_numOTU_data, cycles == "15x", taq == "Phusion") %>% pull(taq_min)` - `r filter(fecal_numOTU_data, cycles == "15x", taq == "Phusion") %>% pull(taq_max)` versus Phusion at 35 cycles = `r filter(fecal_numOTU_data, cycles == "35x", taq == "Phusion") %>% pull(taq_min)` - `r filter(fecal_numOTU_data, cycles == "35x", taq == "Phusion") %>% pull(taq_max)`) [Figure 3B]. Unlike mock community samples, an overall difference in the number of OTUs was only detected between polymerases at 35 cycles in fecal samples (F-stat > `r round(min(filter(fs_anova_summary, cycle == "35x", sub_sample_level == 1000)[, "F.value"]), digits = 2)`, P-value = `r format(filter(fs_anova_summary, cycle == "35x", sub_sample_level == 1000)[, "bh"], digits = 2, nsmall = 2, scientific = T)`) [Table S4], and a Tukey post-hoc test failed to identify any differences between specific polymerase groups (P-value `r ifelse(min(filter(fs_tukey_summary, sub_sample_level == 1000)[, "padj"]) > 0.05,  invisible(">"), invisible("check data"))` 0.05) [Table S5]. Overall, mock communities showed a noticeably larger difference between polymerases in the number of OTUs than actual fecal samples [Figure 3 & Table S4-S5]. This was consistent across the number of amplification cycles tested [Figure 3]. Based on these observations, it is clear that larger differences in total number of OTUs based on polymerases are found in mock versus fecal communities.

With respect to fecal samples, within each respective 5-cycle increment comparison there was no difference in Bray-Curtis index between the polymerases (P-value `r ifelse(min(filter(bray_sim_to_prev_cycle_kruskal_results, sub_sample_level == 1000)[, "bh"], na.rm = T) > 0.05, invisible(">"), invisible("check data"))` 0.05) [Figure 4B]. Using PERMANOVA to test for community differences based on any of the number of cycles within polymerases, only Phusion had cycle dependent differences (P-value = `r format(filter(bray_permanova_by_taq_results, sub_sample_level == 1000, taq == "PHU")[, "bh"], digits = 1, nsmall = 1)`). Additionally, Phusion was one of two polymerases that had enough sequences to be rarefied to 1000 sequences at 15 cycles. There was an overall decrease in Bray-Curtis index by cycle comparison group (e.g. the 15 cycle versus 20 cycle group compared to the 30 cycle versus 35 cycle group) (P-value `r ifelse(pull(fecal_kruskal_test_by_cycle_compare_summary, p.value) < 0.01, invisible("<"), invisible(">"))` 0.01) [Figure 4B]. Using a Dunn's post-hoc test the 15 cycle versus 20 cycle and 20 cycle versus 25 cycle comparison groups had a higher Bray-Curtis index than the 30 cycle versus 35 cycle comparison group (P-value `r ifelse(min(pull(dunns_test_by_cycle_compare_summary, P.adj)) < 0.05, invisible("<"), invisible(">"))` 0.05). Overall, these data suggest that the number of cycles can change the bacterial community independent of the polymerase used to generate the sequences.

When differences in OTUs based on polymerase were investigated for fecal samples, no differences could be found at any number of cycles (P-value $\geqslant$ `r round(min(pull(fecal_otu_tests, p.value)), digits = 2)`, corrected P-value = `r round(min(pull(fecal_otu_tests, bh)), digits = 2)`). These results suggest that taxa specific differences in mock community samples do not carry over to actual fecal community samples.

Using 16S rRNA gene sequencing data there was a large disparity in a model's ability to correctly classify samples based on polymerase used, number of cycles, and individual [Figure 5A]. The model classification success for an individual was significantly better than all other models tested [Figure 5A & Table S8]. Additionally, the top 10 most important variables to the polymerase and number of cycle models were not the most important variables in the model classifying individuals [Figure 5B]. This observation shows that OTUs that change due to the number of cycles and polymerase used are not the same as OTUs that differentiate between individuals [Figure 5B]. Although there are a variety of errors associated with the amplification process, they are smaller than the bacterial community variation between individuals.

>>>>

fecal_sample_rf_summary <- read_csv("../data/process/tables/sample_group_rf_model_summary.csv") 
fecal_poly_rf_summary <- read_csv("../data/process/tables/polymerase_group_rf_model_summary.csv")
fecal_cycles_rf_summary <- read_csv("../data/process/tables/cycles_group_rf_model_summary.csv")

@@

The median error rate varied by polymerase, with Kappa having the highest error rate of all the polymerases across the number of cycles [Figure 2A & Table S2]. Thus, the majority of the differences observed across the number of cycles was between Kappa and other polymerases [Figure 2A & Table S3]. The total sequences that contained at least one error was also polymerase dependent, with the majority of differences being between Kappa or Accuprime and the other polymerases [Table S2 & S3]. These differences in error rates were not due to polymerase dependent differences in base substitution rate [Figure S1]. Collectively, the results suggest that sequence error is dependent on polymerase, persists across the number of cycles used, and are not due to any bias towards a specific type of base substitution.

We next examined whether chimeras also were dependent on polymerase and whether this could affect the total number of OTUs. We observed significant differences in the chimera prevalence based on polymerase across the number of cycles used (P-value `r ifelse(max(filter(kruskal_chimera_prev)[, "bh"]) < 0.05, invisible("<"), invisible("check data"))` 0.05) [Table S2]. Differences in chimera prevalence between Platinum and all other polymerases accounted for the majority of these differences [Table S3]. Accuprime^TM^ had the lowest chimera prevalence of all polymerases regardless of whether chimera removal with VSEARCH was used [Figure 2B & 2C]. The number of cycles used clearly increased chimera prevalence for all polymerases but the rate of increase differed [Figure 2B & 2C]. Additionally, there was a plateau in the total percent of chimeras that were removed that was similar for all polymerases [Figure 2D]. A positive correlation was observed between chimeric sequences and the number of OTUs for all polymerases [Figure 2E]. This positive correlation was strongest for Accuprime^TM^, Platinum, and Phusion [Figure 2E]. This data suggests that chimera prevalence depends on polymerase, is made worse by increasing the number of cycles, and confirms previous reports that the number of OTUs is dependent on the prevalence of these chimeric sequences.

@@

We next investigated the effect polymerases and number of cycles had on the bacterial community composition in more detail by using the Bray-Curtis index. We observed no difference in the bacterial community composition between any measure tested when using the mock community (P-value `r ifelse(pull(mock_kruskal_test_by_cycle_compare_summary, p.value) < 0.05, invisible("<"), invisible(">"))` 0.05) [Figure 4A]. Finding minimal differences in overall bacterial community composition, we next examined if specific taxa vary depending on polymerase and number of cycles. We found that OTUs that taxonomically classified to *Salmonella*, *Escherichia*, *Enterococcus*, and *Staphylococcus* differed based on polymerase used [Table S6 & S7]. These specific OTUs were all members of the mock community and the differences were most pronounced at 25 and 30 cycles [Table S7]. The majority of the differences between taxa were due to differences between a single polymerase (Kappa) and the other polymerases tested [Table S7].

Based on these observations, we created Random Forest models to classify samples based on number of cycles or polymerase used. When assessing the Random Forest models there was a difference between models built to classify number of cycles versus those built to classify polymerase used [P-value = `r format(filter(dunn_rf_data, str_detect(var1, "mock") & str_detect(var2, "mock")) %>% pull(bh), scientific = T, digits = 3)`]. Although there was a difference, neither model was successful at correctly classifying samples based on 16S rRNA gene sequencing data (cycles model, probability of correct classification = `r round(exp(-median(mock_cycles_rf_summary$logLoss)), digits = 2)` (`r round(exp(-max(mock_cycles_rf_summary$logLoss)), digits = 2)` - `r round(exp(-min(mock_cycles_rf_summary$logLoss)), digits = 2)`)(min - max) versus polymerase model, probability of correct classification = `r round(exp(-median(mock_poly_rf_summary$logLoss)), digits = 2)` (`r round(exp(-max(mock_poly_rf_summary$logLoss)), digits = 2)` - `r round(exp(-min(mock_poly_rf_summary$logLoss)), digits = 2)`)). Our observations in mock communities suggest that although differences are detected based on number of cycles or polymerase used, the overall community-wide differences are subtle.

***Few differences based on number of cycles and polymerase used were identified in fecal samples.***
Unlike mock community samples, an overall difference in the number of OTUs was only detected between polymerases at 35 cycles in fecal samples (F-stat > `r round(min(filter(fs_anova_summary, cycle == "35x", sub_sample_level == 1000)[, "F.value"]), digits = 2)`, P-value = `r format(filter(fs_anova_summary, cycle == "35x", sub_sample_level == 1000)[, "bh"], digits = 2, nsmall = 2, scientific = T)`) [Table S4], and a Tukey post-hoc test failed to identify any differences between specific polymerases used (P-value `r ifelse(min(filter(fs_tukey_summary, sub_sample_level == 1000)[, "padj"]) > 0.05,  invisible(">"), invisible("check data"))` 0.05) [Table S5]. However, differences in the range of the number of OTUs detected for fecal samples was dependent on the polymerase used (e.g. Accuprime at 35 cycles = `r filter(fecal_numOTU_data, cycles == "35x", taq == "Accuprime") %>% pull(taq_min)` - `r filter(fecal_numOTU_data, cycles == "35x", taq == "Accuprime") %>% pull(taq_max)` versus Phusion at 35 cycles = `r filter(fecal_numOTU_data, cycles == "35x", taq == "Phusion") %>% pull(taq_min)` - `r filter(fecal_numOTU_data, cycles == "35x", taq == "Phusion") %>% pull(taq_max)`) [Figure 3B]. There was also a trend for lower number of cycles (15-20) to result in a reduced range in the number of OTUs versus higher number of cycles (25, 30, and 35) for all polymerases (e.g. Phusion at 15 cycles = `r filter(fecal_numOTU_data, cycles == "15x", taq == "Phusion") %>% pull(taq_min)` - `r filter(fecal_numOTU_data, cycles == "15x", taq == "Phusion") %>% pull(taq_max)` versus Phusion at 35 cycles = `r filter(fecal_numOTU_data, cycles == "35x", taq == "Phusion") %>% pull(taq_min)` - `r filter(fecal_numOTU_data, cycles == "35x", taq == "Phusion") %>% pull(taq_max)`) [Figure 3B]. Based on these observations, there are only small differences in the total number of OTUs detected in fecal samples that are based on number of cycles or polymerases used.

When using the Bray-Curtis index, we observed that the affect polymerases and number of cycles had on the bacterial community composition in fecal samples was minimal. Within each respective 5-cycle increment comparison there was no difference in Bray-Curtis index between the polymerases (P-value `r ifelse(min(filter(bray_sim_to_prev_cycle_kruskal_results, sub_sample_level == 1000)[, "bh"], na.rm = T) > 0.05, invisible(">"), invisible("check data"))` 0.05) [Figure 4B]. When using PERMANOVA to test for community differences based on any of the number of cycles within polymerases, only Phusion had cycle dependent differences (P-value = `r format(filter(bray_permanova_by_taq_results, sub_sample_level == 1000, taq == "PHU")[, "bh"], digits = 1, nsmall = 1)`). Additionally, Phusion was one of two polymerases that had enough sequences to be rarefied to 1000 sequences at 15 cycles. There was also an overall decrease in Bray-Curtis index by cycle comparison group (e.g. the 15 cycle versus 20 cycle group compared to the 30 cycle versus 35 cycle group) (P-value `r ifelse(pull(fecal_kruskal_test_by_cycle_compare_summary, p.value) < 0.01, invisible("<"), invisible(">"))` 0.01) [Figure 4B]. Using a Dunn's post-hoc test the 15 cycle versus 20 cycle and 20 cycle versus 25 cycle comparison groups had a higher Bray-Curtis index than the 30 cycle versus 35 cycle comparison group (P-value `r ifelse(min(pull(dunns_test_by_cycle_compare_summary, P.adj)) < 0.05, invisible("<"), invisible(">"))` 0.05). In addition to assessing community composition by Bray-Curtis index, we also investigated differences in OTUs based on polymerase at each number of cycle used and no significant differences could be found (P-value $\geqslant$ `r round(min(pull(fecal_otu_tests, p.value)), digits = 2)`, corrected P-value = `r round(min(pull(fecal_otu_tests, bh)), digits = 2)`). Overall, these results suggest that in fecal samples the number of cycles and polymerase used only minimally change the bacterial community composition.

Finally, we built Random Forest models to classify samples for the number of cycles, polymerase used, or individual it was obtained from. Using 16S rRNA gene sequencing data there was a large disparity in the model's ability to correctly classify samples based on polymerase used, number of cycles, and individual (P-value $\leqslant$ `r format(max(filter(dunn_rf_data, str_detect(var1, "fecal") & str_detect(var2, "fecal")) %>% pull(bh)), scientific = T, digits = 3)`). There was a clear difference in the success of the model used to classify individuals (probability of correct classification = `r round(exp(-median(fecal_sample_rf_summary$logLoss)), digits = 2)` (`r round(exp(-max(fecal_sample_rf_summary$logLoss)), digits = 2)` - `r round(exp(-min(fecal_sample_rf_summary$logLoss)), digits = 2)`)) versus the models created to classify number of cycles or polymerase (cycles model, probability of correct classification = `r round(exp(-median(fecal_cycles_rf_summary$logLoss)), digits = 2)` (`r round(exp(-max(fecal_cycles_rf_summary$logLoss)), digits = 2)` - `r round(exp(-min(fecal_cycles_rf_summary$logLoss)), digits = 2)`)(min - max) and polymerase model, probability of correct classification = `r round(exp(-median(fecal_poly_rf_summary$logLoss)), digits = 2)` (`r round(exp(-max(fecal_poly_rf_summary$logLoss)), digits = 2)` - `r round(exp(-min(fecal_poly_rf_summary$logLoss)), digits = 2)`)). Additionally, the top 10 most important OTUs to the polymerase and number of cycle models were not the most important OTUs to the model that classified individuals [Figure 5B]. This suggests that OTUs that change the most due to the number of cycles and polymerase in fecal samples are not the same as OTUs that are part of the biological variation between individuals [Figure 5B]. Collectively, our observations show that despite a variety of errors associated with the amplification process, it is still possible to detect the bacterial community variation between individuals.

### 2018-07-13 submission/manuscript.tex
revamp results section

The median error rate varied by polymerase, with Kappa having the
highest error rate of all the polymerases across the number of cycles
{[}Figure 2A \& Table S2{]}. Thus, the majority of the differences
observed across the number of cycles was between Kappa and other
polymerases {[}Figure 2A \& Table S3{]}. The total sequences that
contained at least one error was also polymerase dependent, with the
majority of differences being between Kappa or Accuprime and the other
polymerases {[}Table S2 \& S3{]}. These differences in error rates were
not due to polymerase dependent differences in base substitution rate
{[}Figure S1{]}. Collectively, the results suggest that sequence error
is dependent on polymerase, persists across the number of cycles used,
and are not due to any bias towards a specific base substitution.

Based on the previous results, we examined whether chimeras also were
dependent on polymerase and whether this could affect the total number
of OTUs. We observed significant differences in the chimera prevalence
based on polymerase across the number of cycles used (P-value
\textless{} 0.05) {[}Table S2{]}. Differences in chimera prevalence
between Platinum and all other polymerases accounted for the majority of
these differences {[}Table S3{]}. Accuprime\textsuperscript{TM} had the
lowest chimera prevalence of all polymerases regardless of whether
chimera removal with VSEARCH was used {[}Figure 2B \& 2C{]}. The number
of cycles used clearly increased chimera prevalence for all polymerases
but the rate of increase differed {[}Figure 2B \& 2C{]}. Additionally,
there was a plateau in the total percent of chimeras that were removed
that was similar for all polymerases {[}Figure 2D{]}. A positive
correlation was observed between chimeric sequences and the number of
OTUs for all polymerases {[}Figure 2E{]}. This positive correlation was
strongest for Accuprime\textsuperscript{TM}, Platinum, and Phusion
{[}Figure 2E{]}. This data suggests that chimera prevalence depends on
polymerase, is made worse by increasing the number of cycles, and
confirms that the number of OTUs is dependent on the prevalence of these
chimeric sequences.

@@

Next, we investigated the effect polymerases and number of cycles had on
the bacterial community composition in more detail by using the
Bray-Curtis index. We observed no difference in the bacterial community
composition between any measure tested when using the mock community
(P-value \textgreater{} 0.05) {[}Figure 4A{]}. Finding minimal
differences in overall bacterial community composition, we next examined
if specific taxa vary depending on polymerase and number of cycles. We
found that OTUs that taxonomically classified to \emph{Salmonella},
\emph{Escherichia}, \emph{Enterococcus}, and \emph{Staphylococcus}
differed based on polymerase used {[}Table S6 \& S7{]}. These specific
OTUs were all members of the mock community and the differences were
most pronounced at 25 and 30 cycles {[}Table S7{]}. The majority of the
differences between taxa were due to differences between a single
polymerase (Kappa) and the other polymerases {[}Table S7{]}.

When assessing the Random Forest models there was a difference between
models built to classify number of cycles versus those built to classify
polymerase used {[}P-value = 5.06e-07{]}. Although there was a
difference, neither model was successful at correctly classifying
samples based on 16S rRNA gene sequencing data (cycles model,
probability of correct classification = 0.42 (0.4 - 0.43)(min - max)
versus polymerase model, probability of correct classification = 0.39
(0.37 - 0.4)). Our observations in mock communities suggest that
although differences are detected based on number of cycles or
polymerase used, the overall community-wide differences are subtle.

\textbf{\emph{Few differences based on number of cycles and polymerased
used were identified in fecal samples.}} Similarly, differences in the
range of the number of OTUs detected for fecal samples was dependent on
the polymerase used (e.g.~Accuprime at 35 cycles = 84 - 106 versus
Phusion at 35 cycles = 84 - 136) {[}Figure 3B{]}. There was also a trend
for lower number of cycles (15-20) to result in a reduced range in the
number of OTUs versus higher number of cycles (25, 30, and 35) for all
polymerases (e.g.~Phusion at 15 cycles = 10 - 19 versus Phusion at 35
cycles = 84 - 136) {[}Figure 3B{]}. Unlike mock community samples, an
overall difference in the number of OTUs was only detected between
polymerases at 35 cycles in fecal samples (F-stat \textgreater{} 16.35,
P-value = 9.7e-05) {[}Table S4{]}, and a Tukey post-hoc test failed to
identify any differences between specific polymerase groups (P-value
\textgreater{} 0.05) {[}Table S5{]}. Overall, mock communities showed a
noticeably larger difference between polymerases in the number of OTUs
than actual fecal samples {[}Figure 3 \& Table S4-S5{]}. This was
consistent across the number of amplification cycles tested {[}Figure
3{]}. Based on these observations, it is clear that larger differences
in total number of OTUs based on polymerases are found in mock versus
fecal communities.

With respect to fecal samples, within each respective 5-cycle increment
comparison there was no difference in Bray-Curtis index between the
polymerases (P-value \textgreater{} 0.05) {[}Figure 4B{]}. Using
PERMANOVA to test for community differences based on any of the number
of cycles within polymerases, only Phusion had cycle dependent
differences (P-value = 0.03). Additionally, Phusion was one of two
polymerases that had enough sequences to be rarefied to 1000 sequences
at 15 cycles. There was an overall decrease in Bray-Curtis index by
cycle comparison group (e.g.~the 15 cycle versus 20 cycle group compared
to the 30 cycle versus 35 cycle group) (P-value \textless{} 0.01)
{[}Figure 4B{]}. Using a Dunn's post-hoc test the 15 cycle versus 20
cycle and 20 cycle versus 25 cycle comparison groups had a higher
Bray-Curtis index than the 30 cycle versus 35 cycle comparison group
(P-value \textless{} 0.05). Overall, these data suggest that the number
of cycles can change the bacterial community independent of the
polymerase used to generate the sequences.

When differences in OTUs based on polymerase were investigated for fecal
samples, no differences could be found at any number of cycles (P-value
\(\geqslant\) 0.01, corrected P-value = 1). These results suggest that
taxa specific differences in mock community samples do not carry over to
actual fecal community samples.

Using 16S rRNA gene sequencing data there was a large disparity in a
model's ability to correctly classify samples based on polymerase used,
number of cycles, and individual {[}Figure 5A{]}. The model
classification success for an individual was significantly better than
all other models tested {[}Figure 5A \& Table S8{]}. Additionally, the
top 10 most important variables to the polymerase and number of cycle
models were not the most important variables in the model classifying
individuals {[}Figure 5B{]}. This observation shows that OTUs that
change due to the number of cycles and polymerase used are not the same
as OTUs that differentiate between individuals {[}Figure 5B{]}. Although
there are a variety of errors associated with the amplification process,
they are smaller than the bacterial community variation between
individuals.

>>>>

The median error rate varied by polymerase, with Kappa having the
highest error rate of all the polymerases across the number of cycles
{[}Figure 2A \& Table S2{]}. Thus, the majority of the differences
observed across the number of cycles was between Kappa and other
polymerases {[}Figure 2A \& Table S3{]}. The total sequences that
contained at least one error was also polymerase dependent, with the
majority of differences being between Kappa or Accuprime and the other
polymerases {[}Table S2 \& S3{]}. These differences in error rates were
not due to polymerase dependent differences in base substitution rate
{[}Figure S1{]}. Collectively, the results suggest that sequence error
is dependent on polymerase, persists across the number of cycles used,
and are not due to any bias towards a specific type of base
substitution.

We next examined whether chimeras also were dependent on polymerase and
whether this could affect the total number of OTUs. We observed
significant differences in the chimera prevalence based on polymerase
across the number of cycles used (P-value \textless{} 0.05) {[}Table
S2{]}. Differences in chimera prevalence between Platinum and all other
polymerases accounted for the majority of these differences {[}Table
S3{]}. Accuprime\textsuperscript{TM} had the lowest chimera prevalence
of all polymerases regardless of whether chimera removal with VSEARCH
was used {[}Figure 2B \& 2C{]}. The number of cycles used clearly
increased chimera prevalence for all polymerases but the rate of
increase differed {[}Figure 2B \& 2C{]}. Additionally, there was a
plateau in the total percent of chimeras that were removed that was
similar for all polymerases {[}Figure 2D{]}. A positive correlation was
observed between chimeric sequences and the number of OTUs for all
polymerases {[}Figure 2E{]}. This positive correlation was strongest for
Accuprime\textsuperscript{TM}, Platinum, and Phusion {[}Figure 2E{]}.
This data suggests that chimera prevalence depends on polymerase, is
made worse by increasing the number of cycles, and confirms previous
reports that the number of OTUs is dependent on the prevalence of these
chimeric sequences.

@@

We next investigated the effect polymerases and number of cycles had on
the bacterial community composition in more detail by using the
Bray-Curtis index. We observed no difference in the bacterial community
composition between any measure tested when using the mock community
(P-value \textgreater{} 0.05) {[}Figure 4A{]}. Finding minimal
differences in overall bacterial community composition, we next examined
if specific taxa vary depending on polymerase and number of cycles. We
found that OTUs that taxonomically classified to \emph{Salmonella},
\emph{Escherichia}, \emph{Enterococcus}, and \emph{Staphylococcus}
differed based on polymerase used {[}Table S6 \& S7{]}. These specific
OTUs were all members of the mock community and the differences were
most pronounced at 25 and 30 cycles {[}Table S7{]}. The majority of the
differences between taxa were due to differences between a single
polymerase (Kappa) and the other polymerases tested {[}Table S7{]}.

Based on these observations, we created Random Forest models to classify
samples based on number of cycles or polymerase used. When assessing the
Random Forest models there was a difference between models built to
classify number of cycles versus those built to classify polymerase used
{[}P-value = 5.06e-07{]}. Although there was a difference, neither model
was successful at correctly classifying samples based on 16S rRNA gene
sequencing data (cycles model, probability of correct classification =
0.42 (0.4 - 0.43)(min - max) versus polymerase model, probability of
correct classification = 0.39 (0.37 - 0.4)). Our observations in mock
communities suggest that although differences are detected based on
number of cycles or polymerase used, the overall community-wide
differences are subtle.

\textbf{\emph{Few differences based on number of cycles and polymerase
used were identified in fecal samples.}} Unlike mock community samples,
an overall difference in the number of OTUs was only detected between
polymerases at 35 cycles in fecal samples (F-stat \textgreater{} 16.35,
P-value = 9.7e-05) {[}Table S4{]}, and a Tukey post-hoc test failed to
identify any differences between specific polymerases used (P-value
\textgreater{} 0.05) {[}Table S5{]}. However, differences in the range
of the number of OTUs detected for fecal samples was dependent on the
polymerase used (e.g.~Accuprime at 35 cycles = 84 - 106 versus Phusion
at 35 cycles = 84 - 136) {[}Figure 3B{]}. There was also a trend for
lower number of cycles (15-20) to result in a reduced range in the
number of OTUs versus higher number of cycles (25, 30, and 35) for all
polymerases (e.g.~Phusion at 15 cycles = 10 - 19 versus Phusion at 35
cycles = 84 - 136) {[}Figure 3B{]}. Based on these observations, there
are only small differences in the total number of OTUs detected in fecal
samples that are based on number of cycles or polymerases used.

When using the Bray-Curtis index, we observed that the affect
polymerases and number of cycles had on the bacterial community
composition in fecal samples was minimal. Within each respective 5-cycle
increment comparison there was no difference in Bray-Curtis index
between the polymerases (P-value \textgreater{} 0.05) {[}Figure 4B{]}.
When using PERMANOVA to test for community differences based on any of
the number of cycles within polymerases, only Phusion had cycle
dependent differences (P-value = 0.03). Additionally, Phusion was one of
two polymerases that had enough sequences to be rarefied to 1000
sequences at 15 cycles. There was also an overall decrease in
Bray-Curtis index by cycle comparison group (e.g.~the 15 cycle versus 20
cycle group compared to the 30 cycle versus 35 cycle group) (P-value
\textless{} 0.01) {[}Figure 4B{]}. Using a Dunn's post-hoc test the 15
cycle versus 20 cycle and 20 cycle versus 25 cycle comparison groups had
a higher Bray-Curtis index than the 30 cycle versus 35 cycle comparison
group (P-value \textless{} 0.05). In addition to assessing community
composition by Bray-Curtis index, we also investigated differences in
OTUs based on polymerase at each number of cycle used and no significant
differences could be found (P-value \(\geqslant\) 0.01, corrected
P-value = 1). Overall, these results suggest that in fecal samples the
number of cycles and polymerase used only minimally change the bacterial
community composition.

Finally, we built Random Forest models to classify samples for the
number of cycles, polymerase used, or individual it was obtained from.
Using 16S rRNA gene sequencing data there was a large disparity in the
model's ability to correctly classify samples based on polymerase used,
number of cycles, and individual (P-value \(\leqslant\) 5.49e-07). There
was a clear difference in the success of the model used to classify
individuals (probability of correct classification = 0.87 (0.86 - 0.9))
versus the models created to classify number of cycles or polymerase
(cycles model, probability of correct classification = 0.26 (0.25 -
0.27)(min - max) and polymerase model, probability of correct
classification = 0.18 (0.17 - 0.2)). Additionally, the top 10 most
important OTUs to the polymerase and number of cycle models were not the
most important OTUs to the model that classified individuals {[}Figure
5B{]}. This suggests that OTUs that change the most due to the number of
cycles and polymerase in fecal samples are not the same as OTUs that are
part of the biological variation between individuals {[}Figure 5B{]}.
Collectively, our observations show that despite a variety of errors
associated with the amplification process, it is still possible to
detect the bacterial community variation between individuals.
